---
title: "Advanced Statistical Computing - Exam"
author: "Justin Kraaijenbrink"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

### Name: Justin Kraaijenbrink

### ULCN: s2577984

```{r label = preps, echo=FALSE}
rm(list=ls()) #clears your workspace
```

Load packages:
```{r}
library(ggplot2)
library(ggpmisc)
library(tidyverse)
library(copula)
library(gridExtra)
```

Read in data:
```{r}
data <- read.csv("insurance.csv") %>% 
  select(2:3)
attach(data)
```

## Task 1. Dependence in the data
Correlation:
```{r}
rho1 <- cor(PLI, WC)
rho2 <- cor(PLI[PLI < max(PLI)], WC[PLI < max(PLI)])
```

Histograms of PLI and WC:
```{r}
pdf("images/hists.pdf")
ggplot(data, aes(x = PLI)) +
  geom_histogram(aes(fill = "PLI"), binwidth = 1, alpha = 0.5) +
  geom_histogram(aes(x = WC, fill = "WC"), binwidth = 1, alpha = 0.5) +
 scale_fill_manual(name = "Business line",
                        breaks = c("PLI", "WC"),
                        values = c("PLI" = "navy", "WC" = "orange"))
dev.off()
```

Scatterplot (joint distribution):
```{r}
pdf("images/joint.pdf")
hist_top <- ggplot(data, aes(x = PLI)) +
  geom_histogram(aes(y = ..density..), colour = 'black', fill = 'grey', binwidth = 1) +
  geom_density(alpha = .2, fill = "blue")

hist_side <- ggplot(data, aes(x = WC)) +
  geom_histogram(aes(y = ..density..), colour = 'black', fill = 'grey', binwidth = 1) +
  geom_density(alpha = .2, fill = "blue") +
  coord_flip()

scatter <- ggplot(data, aes(PLI, WC)) +
  geom_point() +
  geom_point(aes(x = max(PLI), y = max(WC)), colour = 'red') +
  geom_density_2d()

empty <- ggplot() + 
  geom_point(aes(1, 1), colour="white") +
  theme(axis.ticks=element_blank(), 
        panel.background=element_blank(), 
        axis.text.x=element_blank(), 
        axis.text.y=element_blank(),           
        axis.title.x=element_blank(), 
        axis.title.y=element_blank())

grid.arrange(hist_top, empty, scatter, hist_side, 
             ncol = 2, widths = c(4, 1), heights = c(1, 4))
dev.off()
```

## Tasks 3-5. Maximum Likelihood for estimation of parameters
```{r}
loglik <- function(pars, x) -sum(dlnorm(x, pars[1], pars[2], log = TRUE))

loglikJoe <- function(theta, x) -sum(dCopula(x, joeCopula(theta), log = TRUE))

EstimateParams <- function(data) {
  mu0_1 <- mean(log(data[, 1]))
  sd0_1 <- sqrt(sum((log(data[, 1]) - mu0_1)**2)/nrow(data))
  pars_1 <- c(mu0_1, sd0_1)

  mu0_2 <- mean(log(data[, 2]))
  sd0_2 <- sqrt(sum((log(data[, 2]) - mu0_2)**2)/nrow(data))
  pars_2 <- c(mu0_2, sd0_2)
  
  MLE_1 <- optim(pars_1, loglik, x = data[, 1])
  MLE_2 <- optim(pars_2, loglik, x = data[, 2])
  
  U <- cbind(plnorm(data[, 1], MLE_1$par[1], MLE_1$par[2]),
             plnorm(data[, 2], MLE_2$par[1], MLE_2$par[2]))
  
  theta <- optim(1, loglikJoe, x = U, method = "Brent", 
                 lower = 1, upper = 10)$par

  params <- c(MLE_1$par, MLE_2$par, theta)
  
  return(params)
}

params <- EstimateParams(data)
```

# Tasks 6-7. Simulate data
``` {r}
SimulateJoint <- function(n, params) {
  Us <- rCopula(n, joeCopula(params[5]))
  PLI_hat <- qlnorm(Us[, 1], params[1], params[2])
  WC_hat <- qlnorm(Us[, 2], params[3], params[4])
  
  return(data.frame(X1 = PLI_hat, 
                    X2 = WC_hat))
}

n <- nrow(data)
simdata <- SimulateJoint(n, params)

plotdata <- cbind(simdata, PLI, WC)

pdf("images/simdata.pdf")
ggplot(plotdata) + 
  geom_point(aes(x = PLI, y = WC, color = "True"), alpha = 0.5) +
  geom_point(aes(x = X1, y = X2, color = "Simulated"), alpha = 0.5)
dev.off()
```

See how parameters effect simulation:
```{r}
simdata_mu1 <- SimulateJoint(n, c(2 * params[1], params[2:5])) 
simdata_sd1 <- SimulateJoint(n, c(params[1], 2 * params[2], params[3:5]))
simdata_theta <- SimulateJoint(n, c(params[1:4], 4 * params[5]))

plotdata <- data.frame(simdata_mu1, simdata_sd1, simdata_theta, data)

pdf("images/simparams.pdf")
ggplot(plotdata) +
  geom_point(aes(x = X1, y = X2, color = "mu1"), alpha = 0.4) +
  geom_point(aes(x = X1.1, y = X2.1, color = "sd1"), alpha = 0.4) +
  geom_point(aes(x = X1.2, y = X2.2, color = "theta"), alpha = 0.3) +
  geom_point(aes(x = PLI, y = WC, color = "True"), alpha = 0.3)
dev.off()  
```


## Task 8. Simulation study 
```{r}
mu1 <- 1; sigma1 <- 2; mu2 <- 3; sigma2 <- 0.5; theta <- 2
fixed_params <- c(mu1, sigma1, mu2, sigma2, theta)
Ns <- c(100, 200, 500, 1000)
rep <- 100

RMSEs <- matrix(nrow = length(Ns), ncol = length(fixed_params))
computing_time <- numeric(length(Ns))

for (N in 1:length(Ns)) {
  computing_time[N] <- system.time({
    n <- Ns[N]
    par_est <- matrix(nrow = rep, ncol = length(fixed_params))
    for(i in 1:rep) {
      simdata <- SimulateJoint(n, fixed_params) 
      par_est[i, ] <- EstimateParams(simdata)
    }
    
    RMSEs[N, ] <- apply(par_est, 2, sd) / sqrt(rep)
    })[3]
}

RMSEs <- RMSEs %>% 
  `colnames<-` (c("mu1", "sigma1", "mu2", "sigma2", "theta")) %>% 
  `rownames<-` (paste0("N = ", Ns)) %>% 
  as.data.frame

pdf("images/RMSEs.pdf")
ggplot(RMSEs, aes(x = 1:4)) +
  geom_line(aes(y = mu1, color = "mu1")) + 
  geom_line(aes(y = sigma1, color = "sigma1")) +
  geom_line(aes(y = mu2, color = "mu2")) + 
  geom_line(aes(y = sigma2, color = "sigma2")) +
  geom_line(aes(y = theta, color = "theta")) +
  scale_x_continuous(breaks = 1:4, labels = rownames(RMSEs)) +
  ylab("RMSE") +
  xlab("")
dev.off()


pdf("images/computingtimes.pdf", width = 16)
ggplot(data.frame(t = computing_time, n = Ns, N = as.factor(Ns))) +
  geom_bar(aes(x = N, y = t/n, fill = N), stat = "identity") + 
  ylab("Computing time / N") + 
  theme(axis.title.y = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.position = c(0.9, 0.85))
dev.off()
```

## Task 9. Expected pay-out (plain MC)
```{r}
CalculateVt <- function(data, t) {
  V <- rowSums(data) * (rowSums(data) > t)
  
  return(V)
}

MC_plain <- function(N, params, t) {
  X <- SimulateJoint(N, params)
  out <- mean(CalculateVt(X, t))
  
  return(out)
}

params <- EstimateParams(data)
N <- 10^5
Ts <- seq(100, 200, 10)

Vt_plain <- sapply(Ts, MC_plain, N = N, params = params)

pdf("images/VtplainMC.pdf")
ggplot(tibble(t = Ts, Vt = Vt_plain), aes(t, Vt)) +
  geom_point() + 
  scale_x_continuous(breaks = Ts, labels = Ts) +
  geom_smooth(method = "loess", formula = "y ~ x") 
dev.off()
```

## Task 10. Expected pay-out (importance sampling)
```{r}
MC_IS <- function(N, params, Ts) {
  new_params <- c(params[1] + 1, params[2:5])
  Y <- SimulateJoint(N, new_params)
  weights <- dlnorm(Y[, 1], params[1], params[2]) / dlnorm(Y[, 1], new_params[1], new_params[2])
  
  Vt_IS <- numeric(length(Ts))
  
  for(i in 1:length(Ts)) {
    t <- Ts[i]
    Vt_IS[i] <- mean(CalculateVt(Y, t) * weights)
  }
  
  return(Vt_IS)
}

Vt_IS <- MC_IS(N, params, Ts)
P <- 40000*exp(-Ts/7)

plotdata <- data.frame(t = Ts, Vt = Vt_IS, P = P)

pdf("images/VtvsP.pdf", width = 10)
ggplot(plotdata, aes(x = t)) +
  geom_point(aes(y = Vt, color = "Vt"), na.rm = TRUE) + 
  geom_line(aes(y = Vt, color = "Vt"), na.rm = TRUE) + 
  geom_point(aes(y = P, color = "P"), na.rm = TRUE) + 
  geom_line(aes(y = P, color = "P"), na.rm = TRUE) +
  scale_x_continuous(breaks = Ts, labels = Ts) +
  ylim(0, 0.0015) 
dev.off()
```

## Task 11. Bootstrap results
```{r}
BootstrapIS <- function(B, N, data, Ts) {
  
  out <- replicate(B, expr = {
              b_data <- as.data.frame(data[sample(nrow(data), replace = TRUE), ])
              params <- EstimateParams(b_data)
              Vt_b <- MC_IS(N, params, Ts)})
  
  return(out)
}

n <- 648
N <- 10^5
B <- 1000

Vt_boot <- BootstrapIS(B, N, data, Ts)

quantiles <- apply(Vt_boot, 1, quantile, c(0.10, 0.90))
bounds <- rbind(2 * rowMeans(Vt_boot) - quantiles[2, ],
                2 * rowMeans(Vt_boot) + quantiles[1, ])

plotdata <- data.frame(plotdata,
                       lower = bounds[1, ],
                       upper = bounds[2, ])

pdf("images/Vtboot.pdf", width = 10)
ggplot(plotdata, aes(x = t)) +
  geom_point(aes(y = Vt, color = "Vt"), na.rm = TRUE) + 
  geom_line(aes(y = Vt, color = "Vt"), na.rm = TRUE) + 
  geom_errorbar(aes(ymin = lower, ymax = upper, color = "Vt"), width = 1) +
  geom_point(aes(y = P, color = "P"), na.rm = TRUE) + 
  geom_line(aes(y = P, color = "P"), na.rm = TRUE) +
  scale_x_continuous(breaks = Ts, labels = Ts) +
  ylim(-1e-5, 0.0017)
dev.off()
```

```{r}
save.image()
```

